---
title: "Modeling code"
output: github_document
date: 2020-03-30
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
```

## Initial Parameters

```{r}
## BIG TODO: add citations for these numbers

initial_population_size = 20 # TODO - choose bigger number later
hospital_bed_capacity = 100 #TODO - fill in with real number later

probability_symptomatic = 2/3
probability_asymptomatic = 1 - probability_symptomatic

# TODO: break this out into it's own code chunk with explanation? Weighted average, solved equations to get 
basic_reproductive_number = 2.4
symptomatic_reproductive_number = basic_reproductive_number*(6/5)
asymptomatic_reproductive_number = basic_reproductive_number*(3/5)

time_until_infectious = 4.5 # This is figured from the incubation period, and rounded down from 4.6 so I can do steps in half-days.

# For the asymptomatic types
asymptomatic_time_from_infectious_to_recovery = 4 # This is figured from the generation time (confusing). Should this be longer??

# For the symptomatic types
time_infectious_before_symptom_onset = 0.5
time_from_symptom_onset_to_recovery = 3.5

time_until_seeking_care = 5 # This is the number of days after symptom onset that people seek hospital care if they need it, on average
time_in_hospital_bed = 10.5
time_until_death_if_no_care = 10

```

## States

```{r}
SUCCEPTIBLE = "succeptible"

INFECTED_ASYMPTOMATIC = "infected_asymptomatic"
INFECTIOUS_ASYMPTOMATIC = "infectious_asymptomatic"

INFECTED_SYMPTOMATIC_PRE_SYMPTOMS = "infected_symptomatic_pre_symptoms"
INFECTIOUS_SYMPTOMATIC_PRE_SYMPTOMS = "infectious_symptomatic_pre_symptoms"

SYMPTOMATIC_NEED_HOSPITAL = "symptomatic_need_hospital"
SYMPTOMATIC_DONT_NEED_HOSPITAL = "symptomatic_dont_need_hospital"

NEED_HOSPITAL_SEEK_CARE = "need_hospital_seek_care"
NEED_HOSPITAL_DONT_SEEK_CARE = "need_hospital_dont_seek_care"

GET_HOSPITAL_CARE = "get_hospital_care"
DONT_GET_NEEDED_CARE = "dont_get_needed_care"

RECOVERED = "recovered"
DEAD = "dead"
```

## Change states on each timestep

```{r}
# placeholder function that just returns the same state
change_state = function(prev_state){
  as.character(prev_state)
}
```

## Setting up and populating my dataframe

```{r}
# TODO: create the population according to demographic markers, and randomly assign the infected person.
create_initial_population_with_one_infected = function(size){
  one_infected = as.factor(c(INFECTED_SYMPTOMATIC_PRE_SYMPTOMS))
  others_succeptible = rep(as.factor(c(SUCCEPTIBLE)), size - 1)
  
  fct_c(one_infected, others_succeptible)
}

# remove this later, printing just for debugging
create_initial_population_with_one_infected(initial_population_size)
```

This is a table with each row representing one person in the population. The first few columns include demographic and other information about a person, and all of the columns labeled `day_n` represent that person's disease state at time n.
```{r}
# TODO: make a function to generate new day columns based on some initial parameter.
population = tibble(
  person_ids = 1:initial_population_size,
  day_1 = create_initial_population_with_one_infected(initial_population_size)
) %>% 
  mutate(., day_2 = as.factor(map_chr(day_1, change_state)))

# remove this later, printing just for debugging
population
```


## Visualization

First, transform the table into a better shape for graphing, by getting the total state counts at each time step.
```{r}
population_to_visualize =
  pivot_longer(
    population,
    cols = starts_with("day_"),
    names_to = "day",
    values_to = "state",
    names_prefix = "day_"
  ) %>% 
  mutate(day = as.numeric(day)) %>% 
  group_by(day, state) %>% 
  summarize(count = n())

## NOTE:
## For transitory states like "infected", the count represents how many people are infected on that day.
## However, for a state like "dead", the count represents how many people total have died up until that point (because once you reach the "dead" state, you stay at that state forever).


# remove this later, printing just for debugging
population_to_visualize
```

Graph the person-count of each state in a different color, with days on the x-axis.
```{r}
ggplot(population_to_visualize, 
       aes(x=day, y=count, color=state)) + 
  geom_line()
```


